Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-02 _tex/index.tex
Built site for gh-pages

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution. The
mode (or the mean) of this distribution, representing the (average)
relative separation, is given by the difference between the modal
discriminal processes of the stimuli \(S_{BA} = S_{B} - S_{A}\).
Meanwhile, the dispersion of the distribution, reflecting the
variability in the relative separation, is calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
the discriminal processes of the stimuli. This correlation quantifies
the dependence of the judge's perception of the trait in one stimulus on
his perception of the same trait in the other.

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}, assuming a correlation of
\(\rho = 0.6\). The figure reveals that, under these conditions, the
judge perceives text B as having significantly higher quality than text
A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
a result, the dichotomous outcome of this comparison almost certainly
favors text B over text A.

@@

Notably, the correlation between the discriminal processes, \(\rho\),
plays a pivotal role in determining comparison outcomes by shaping the
distribution of the discriminal difference between the stimuli.
Specifically, as the correlation increases, reflecting a stronger
dependence of the judge's perception of quality in one stimulus on his
perception of the other, the distribution of the discriminal difference
narrows. This narrowing ultimately impacts the area under the curve that
determines the comparison outcome and, consequently, the conclusions
drawn from the comparison.

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
the comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming
that the stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

As outlined in the previous section, the assumption of \emph{zero
correlation between stimuli} suggests that during a pairwise comparison,
a judge's perception of quality in one text does not influence his
perception of the same trait in another text (see
Figure~\ref{fig-correlation}, \(\rho=0\)). Thurstone attributed this
independence to the cancellation of potential judges' biases, driven by
two opposing and equally weighted effects occurring during the pairwise
comparisons \citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978}
mathematically demonstrated this cancellation using the BTL model under
the assumption of discriminal processes with additive biases. However,
it is easy to imagine at least two scenarios where the zero correlation
assumption almost certainly does not hold: when the pairwise comparison
involves multidimensional, complex traits with heterogeneous stimuli and
when an additional hierarchical structure is relevant to the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between heterogeneous stimuli
due to certain judges' biases that resist cancellation. Research on text
quality indicates that when judges evaluate these traits, they often
rely on various intricate aspects of the stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
aspects, which are likely neither equally weighted nor opposing, may
influence judges' perceptions unevenly. For example, this could occur
when a judge assessing the argumentative quality of a text places
disproportionate emphasis on grammatical accuracy, ultimately favoring
texts with fewer errors but weaker arguments. Ignoring these relevant
additional ``traits'' can result in biases that resist cancellation,
introducing dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II} and ultimately violating the
assumption of zero correlation. While direct evidence for the specific
mechanisms causing biases is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may introduce dependencies
between stimuli, a statistical phenomenon commonly known as clustering
\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
presence of such hierarchical structures in CJ data, the statistical
handling of this extra source of dependency has been inadequate. For
example, when CJ data includes multiple samples of stimuli from the same
individuals, researchers often rely on (average) estimated BTL scores to
conduct subsequent analyses and tests at the individual hierarchical
level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature strongly
advises against ignoring relevant additional traits or overlooking
clustering (grouping) structures that may introduce dependencies between
stimuli. First, ignoring additional aspects pertinent to the trait of
interest can cause a dimensional mismatch, which can lead to an
overestimation of the precision (reliability) of the trait's measurement
\citep[pp.~340-341]{Hoyle_et_al_2023} or, worse, introduce bias into the
measurement process \citep{Ackerman_1989}. Additionally, overlooking
relevant hierarchical (or grouping) structures can exacerbate the
inflation of reliability by reducing the accuracy of the model's
parameters \citep[pp.~482]{Hoyle_et_al_2023}. Taken together, these
oversights undermine the reliability of the measurement process and,
since validity cannot exist without reliability
\citep[pp.~2]{Perron_et_al_2015}, ultimately compromise the validity of
the results derived from it.

>>>>

Indeed, the theory assumes that the observed dichotomous outcome arises
from the distribution of the difference between the underlying
discriminal processes of the stimuli, known as the \emph{discriminal
difference}. Since the individual discriminal processes follow a Normal
distribution, their difference also follows a Normal distribution
\citep{Andrich_1978}. The mode (or the mean) of this distribution,
representing the (average) relative separation, is given by the
difference between the modal discriminal processes of the stimuli
\(S_{BA} = S_{B} - S_{A}\). Meanwhile, the dispersion of the
distribution, reflecting the variability in the relative separation, is
calculated as
\(\sigma_{BA} = \sqrt{\sigma_{B}^{2} + \sigma_{A}^{2} - \rho\sigma_{B}\sigma_{A}}\).
Here, \(\sigma_{B}\) and \(\sigma_{A}\) are the previously defined
discriminal dispersions, while \(\rho\) measures the correlation between
the discriminal processes of the stimuli. This correlation quantifies
the dependence of the judge's perception of the trait in one stimulus on
his perception of the same trait in the other.

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}, assuming a correlation of
\(\rho = 0.6\). The figure reveals that, under these conditions, the
judge perceives text B as having significantly higher quality than text
A, as indicated by the shaded gray area under the curve \(P(B > A)\). As
a result, the dichotomous outcome of this comparison would mostly favor
text B over text A.

@@

Notably, the correlation between the discriminal processes, \(\rho\),
plays a pivotal role in determining comparison outcomes by shaping the
distribution of the discriminal difference between the stimuli.
Specifically, as the correlation increases, reflecting a stronger
dependence of the judge's perception of quality in one stimulus on his
perception of the other, the distribution of the discriminal difference
narrows. This narrowing ultimately impacts the area under the curve that
determines the comparison outcome and, consequently, the conclusions
drawn from this outcome.

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
them to a single judge. Case III introduces the assumption of zero
correlation between stimuli. Case IV builds on this by assuming that the
stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

As outlined in Section~\ref{sec-thurstone_theory}, the assumption of
\emph{zero correlation between stimuli} implies that, during a pairwise
comparison, a judge's perception of quality in one text does not
influence his perception of the same trait in another text. Thurstone
attributed this independence to the cancellation of potential judges'
biases, driven by two opposing and equally weighted effects occurring
during the pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
\citet{Andrich_1978} mathematically demonstrated this cancellation using
the BTL model under the assumption of discriminal processes with
additive biases. However, it is easy to imagine at least two scenarios
where the zero correlation assumption almost certainly does not hold:
when the pairwise comparison involves multidimensional, complex traits
with heterogeneous stimuli and when an additional hierarchical structure
is relevant to the stimuli.

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluating these traits, they often
rely on various intricate characteristics of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, consequently, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
disproportionate emphasis on grammatical accuracy, ultimately favoring
texts with fewer errors but weaker arguments. While direct evidence for
this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures in CJ data,
the statistical handling of this extra source of dependency has been
inadequate. For example, when CJ data includes multiple samples of
stimuli from the same individuals, researchers often rely on (average)
estimated BTL scores to conduct subsequent analyses and tests at the
individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature underscores the
importance of incorporating additional relevant traits and clustering
(grouping) structures into the modeling of CJ data to address potential
dependencies between stimuli. Oversights in these areas can lead to
misleading conclusions about the trait's reliability and ultimately
compromise its validity \citep[pp.~2]{Perron_et_al_2015}. Neglecting
additional traits relevant to the stimuli or the trait of interest, such
as judges' biases, introduces a dimensional mismatch in the modeling
process of the CJ data. This mismatch often results in an overestimation
of the trait's reliability \citep[pp.~341]{Hoyle_et_al_2023} or, worse,
introduces biases into the estimated trait \citep{Ackerman_1989}.
Therefore, these traits must be incorporated into the modeling of CJ
data, as demonstrated by \citet{Andrich_1978} and
\citet{Wainer_et_al_1978}. Similarly, ignoring hierarchical (grouping)
structures reduces the accuracy of model parameter estimates and further
exacerbates the overestimation of reliability
\citep[pp.~482]{Hoyle_et_al_2023}. To address this, any additional
hierarchical structure relevant to the stimuli should also be included
in the model, as in Multilevel Structural Equation models (MSEM)
\citep[chap.~26]{Hoyle_et_al_2023}.
